% Cover letter and point-by-point reviewer responses.
% Reviewer text is verbatim from paper_review.txt; \reply{...} blocks
% are scaffolded as TODO placeholders to be filled in during the revision.

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Bioinformatics}
\end{minipage}
\hfill
\begin{minipage}[b]{5.5in}
  Andrew Kern (on behalf of all authors) \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript titled
``MKado: a toolkit for McDonald-Kreitman tests of natural selection''.
We thank the editor and both reviewers for their thoughtful and
constructive comments, which have substantially strengthened the
manuscript and the underlying software package. The most substantive
changes in this revision are: (i) a new aggregated-genes asymptotic-MK
runtime benchmark addressing Reviewer~1, point~5; (ii) implementation of
the Uricchio et al.~(2019) ``above-$x$'' cumulative SFS for the
asymptotic test, addressing Reviewer~1, point~4; (iii) a polyDFE-style
case-resampling bootstrap CI for asymptotic alpha, addressing the second
half of Reviewer~1, point~5; and (iv) prominent VCF-input and
allele-polarization documentation in both manuscript and docs,
addressing Reviewer~1, points~1--2 and Reviewer~2.

In addition to a Response to Reviewers (in which page/line numbers refer
to the revised manuscript file), we provide a marked-up PDF showing the
differences relative to the original submission.

\vskip 2em

\noindent\hspace{4em}
\begin{minipage}{4.5in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf
  Andrew Kern (on behalf of all authors)
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\reviewersection{1}

\begin{itquote}
McDonald and Kreitman (MK) test is one of the most widely used methods
to test for natural selection. Since its original publication, several
studies have proven its potential for detecting positive selection.
Nowadays, we have several extensions that mitigate the main issue of
the MK test: the presence of slightly deleterious mutations. Here, the
authors present a fast, user-friendly, and well-documented software to
automatize MK tests. The software not only includes multiple MK tests
extensions per gene or aggregated datasets, but also data processing,
multiple testing correction, and visualization features. However, I
have a few minor concerns that I hope the authors can address. Most of
them should be straightforward to test or implement.
\end{itquote}

\reply{We thank the reviewer for the positive assessment and for the
detailed, constructive comments. Each point is addressed in turn below;
the substantive items (especially points~4 and~5) drove the
two largest additions in this revision (cumulative-SFS option and a
new aggregated-genes runtime benchmark with bootstrap CIs).}

%---------- R1.1 ----------
\begin{point}{} % 1.1 -- VCF support visible in docs but not in manuscript
As the authors claim, most of the current software is fragmented. Some
tools can deal with VCF or multi-FASTA data (\texttt{degennotate},
\texttt{fastDFE}, \texttt{iMKT}; \url{https://doi.org/10.1093/molbev/msad270},
\url{https://doi.org/10.1093/molbev/msae070},
\url{https://doi.org/10.1093/nar/gkz372}), some have already implemented
several of the MK tests presented here (\texttt{MKtest.jl};
\url{https://doi.org/10.1093/g3journal/jkae031}), others can polarise
alleles (\texttt{fastDFE}; \url{https://doi.org/10.1093/molbev/msae070}).
\mkado unifies it on a user-friendly Python package. The documentation
shows that \mkado can deal with VCF files similarly to \texttt{fastDFE},
but I cannot see this mentioned in the paper. The authors should
include it in the main text, since handling both VCF and multi-FASTA
files is an important feature.
\end{point}

\reply{
We have added a sentence to the Features and Implementation section
of the Methods \revref describing the \texttt{mkado vcf} subcommand
and its inputs (multi-sample ingroup VCF, single-sample outgroup
VCF, reference FASTA, and GFF3 annotation). The same paragraph now
explicitly contrasts the dual FASTA + VCF input capability of
\mkado{} with FASTA-only tools (\texttt{DnaSP},
\texttt{asymptoticMK}) and VCF-only tools (\texttt{degenotate},
\texttt{fastDFE}).
We have also split the ``Sequence input'' row of the feature
comparison table (Table~\ref{tab:comparison}) into ``Sequence
(FASTA) input'' and ``VCF input'' rows, making the difference in
input flexibility across tools visible at a glance.
}

%---------- R1.2 ----------
\begin{point}{} % 1.2 -- automatic allele polarization vs polarized MK
In addition to describing the Polarised MK Test, the authors should
state in the main text that \mkado uses outgroup sequences to polarize
alleles. As far as I understand, this is done automatically and is
well described in the software documentation. Nonetheless, I would
recommend that the authors clarify the automatic allele polarization
required to run tests such as the asymptotic or imputed MK tests
(where polarization is mandatory or recommended), as distinct from the
polarised MK test, which specifically tests for selection along a
lineage. As I said, this is already well explained in the
documentation, so it should be straightforward to include in the paper
to avoid any misunderstanding.
\end{point}

\reply{
We agree that this distinction was not clearly stated. We have added
a short paragraph in the Methods \revref that (i) names the asymptotic
and imputed MK tests as the analyses that require derived allele
frequencies, (ii) states that \mkado{} polarizes alleles automatically
using the outgroup---the allele matching the outgroup is treated as
ancestral and any non-outgroup allele as derived---and (iii) makes
explicit that this automatic single-outgroup polarization is
conceptually distinct from the \textit{polarized} MK test, which
uses a second outgroup to attribute fixed differences to lineages.
The corresponding documentation page (``Allele Polarization'' in the
asymptotic-MK reference) was already in place; the manuscript text
now mirrors it.
}

%---------- R1.3 ----------
\begin{point}{} % 1.3 -- TG alpha + imputed / FWW combination
When using the Tarone-Greenland Alpha, is the standard MK test
applied? Would it not be beneficial to correct for slightly deleterious
mutations by simply using the imputed or Fay, Wyckoff and Wu frequency
threshold correction?
\end{point}

\reply{
The Tarone-Greenland weighted estimator computes a single
genome-wide $\alpha$ from the per-gene
($D_n$, $D_s$, $P_n$, $P_s$) tuples, weighted inversely by
$P_s + D_s$; it does not internally invoke the standard MK test
per gene or its $p$-value.

For the FWW correction: this is already supported. Both the
standard MK and the Tarone-Greenland estimator accept a derived
allele frequency cutoff via \texttt{-{}-min-freq X}; passing
\texttt{-{}-alpha-tg -{}-min-freq 0.15} (or any other threshold)
applies the FWW frequency-threshold correction within
$\alpha_\text{TG}$ exactly. We have made this explicit in the
documentation so that users wanting the FWW-corrected weighted
estimate can easily find the option.

For combining the imputed MK correction with the
Tarone-Greenland weighting: this would be a novel estimator not
described in the literature, and would require new theory for the
weighting (the imputation alters the $P_n$ count in a way that
changes the variance assumptions underlying the
Tarone-Greenland weights). We are not comfortable shipping an
unvalidated weighted-alpha-of-imputed-counts variant as part of a
methods package without independent statistical analysis. We
mention this explicitly here in case the reviewer wishes to
pursue this combination as a separate methodological contribution.
}

%---------- R1.4 ----------
\begin{point}{} % 1.4 -- Uricchio et al. 2019 above-x SFS option
The authors should include an option to modify the SFS as described in
Uricchio et al.\ (2019) (see supplementary material,
\url{https://doi.org/10.1038/s41559-019-0890-6}):

\begin{itquote}
We note that the original asymptotic-MK approach takes $P_N(x)$ and
$P_S(x)$ as the number of polymorphic sites at frequency $x$ rather
than above $x$, but this approach scales poorly as sample size
increases since most common allele frequencies $x$ have very few
polymorphic sites in large samples. We therefore define $P_N(x)$ and
$P_S(x)$ as stated above since these quantities trivially have the
same asymptote but are less affected by changing sample size.
\end{itquote}

The asymptotic estimation would greatly benefit from this modification,
producing more robust results. The authors can follow
\url{https://github.com/jmurga/MKtest.jl/blob/main/src/polymorphism.jl}
to implement it.
\end{point}

\reply{
We agree this was a useful and natural extension. \mkado{} now
supports both forms of the SFS through the \texttt{-{}-sfs-mode}
flag: \texttt{at} (the original per-bin form of
\citealt{messer2013}, where $P_n(x)$ and $P_s(x)$ count sites at
exactly frequency~$x$) and \texttt{above} (the cumulative inclusive
form of \citealt{uricchio2019}, where $P_n(x)$ and $P_s(x)$ count
sites with derived allele frequency at or above~$x$). The default
remains \texttt{at} for backward compatibility; users wanting the
sample-size-robust behaviour described by \citet{uricchio2019} can
opt in with a single flag. The new option is described in the
Methods \revref and is exercised in the new aggregated-asymptotic
runtime benchmark (\cref{fig:aggregated_runtime}), which sweeps the
full \{at, above\} $\times$ \{monte-carlo, bootstrap\} grid and
shows that the asymptotic $\alpha$ estimates from the two SFS
constructions agree across replicates while the cumulative form
produces slightly tighter per-bin $\alpha(x)$ trajectories at high
frequency, as predicted.
}

%---------- R1.5 ----------
\begin{point}{} % 1.5 -- runtime benchmarks: composition + aggregated
Do the runtime plots show only the batch processing of the 12,437
human genes, or also the execution of the imputed or standard MK test?
If the tests are not included on runtime, I would recommend that the
authors incorporate both processing and MK test estimation, since this
should not substantially affect the benchmark. Moreover, I'm assuming
the author are not providing runtime benchmark for asymptotic
estimations. Because asymptotic estimation would probably fail at the
gene level or provide unreliable results (as the authors properly
note), I will recommend two approaches:

\begin{itemize}
\item[(i)] Following Murga-Moreno et al.\ (2022)
(\url{https://doi.org/10.1093/g3journal/jkac206}), the authors can
aggregate at least 1{,}000 random genes to ensure reliable estimation
on human data and run the benchmark over a reasonable number of
replicates.
\item[(ii)] They can add the polyDFE bootstrapping strategy to resample
the SFS over the 12{,}437 aggregated genes. This is a common strategy
for obtaining reliable alpha confidence interval (CI) estimates (e.g.,
Castellano et al.\ 2019;
\url{https://doi.org/10.1534/genetics.119.302494}). The authors
currently estimate CI via Monte Carlo, but adding a bootstrap
resampling strategy would benefit both runtime benchmarks and the
reliability of alpha estimates for any MK test provided. See the
\texttt{bootstrapData} function at
\url{https://github.com/paula-tataru/polyDFE/blob/master/postprocessing.R}
or
\url{https://fastdfe.readthedocs.io/en/latest/_modules/fastdfe/spectrum.html#Spectrum.resample}.
\end{itemize}

This way, non-expert users would not face runtimes longer than those
shown in Figure S1. Since the software scaling is nearly perfect,
testing approximately 1{,}000 aggregated datasets should be
informative enough.
\end{point}

\reply{
Both points are well taken; we have addressed each, and these
changes drove the largest additions in this revision.

\textit{Composition of the existing benchmark.} Figure~S1 measures
the full \texttt{mkado batch} pipeline end-to-end: FASTA parsing
and ingroup/outgroup partitioning, codon classification and
Nei--Gojobori site counting, the per-gene MK test (standard MK is
the test exercised here), Benjamini--Hochberg FDR correction
across all genes, and TSV serialization of the results. There is
no separate ``parse only'' mode being measured. We have added an
explicit statement of this in the Methods text \revref{} and in
the Figure~S1 caption.

\textit{Aggregated-asymptotic benchmark (option (i)).} We have
added a new benchmark in the regime where asymptotic MK is
statistically reliable, following \citet{murga-moreno2022}:
\textbf{1{,}000 random genes per replicate, 100 replicates per
condition}. To exercise the two new asymptotic-MK options
introduced in this revision, the benchmark sweeps the full
2$\times$2 grid of \{\texttt{sfs\_mode at, above}\}
$\times$ \{\texttt{ci\_method monte-carlo, bootstrap}\}, for 400
timed runs per execution. Per-replicate wall time covers the
\texttt{asymptotic\_mk\_test\_aggregated} call only (aggregation,
curve fit, and CI computation); per-gene polymorphism extraction
is performed once across the full 12{,}437-gene corpus and
amortized across replicates---this composition is recorded
explicitly in the figure caption so that the cost breakdown is
unambiguous. The new figure (\cref{fig:aggregated_runtime}) shows
that aggregated $\alpha$ estimates are consistent across all four
conditions, that the case-resampling bootstrap is roughly
2$\times$ slower than parametric Monte-Carlo per replicate as
expected, and that end-to-end pipeline scaling on this aggregated
workload extends usefully to higher worker counts than the
per-gene benchmark of Figure~S1, with a sweet spot near 32
workers. The benchmark is fully reproducible via
\texttt{examples/benchmarks/aggregated\_asymptotic\_runtime.py}
and is documented at \texttt{examples/benchmarks/README.md}.

\textit{Bootstrap CI (option (ii)).} We have added a
case-resampling bootstrap CI for aggregated asymptotic alpha
(\texttt{-{}-ci-method bootstrap}), implementing the same idea as
the polyDFE \texttt{bootstrapData} routine recommended by the
reviewer: at each replicate, the pooled polymorphism list is
resampled with replacement, the SFS is rebinned, the asymptotic
model is refit, and the asymptotic-$\alpha$ value is recorded; the
CI is read off the resulting empirical distribution. This is
available alongside the original parametric Monte-Carlo CI, and is
selectable via \texttt{-{}-ci-method bootstrap} or
\texttt{-{}-ci-method monte-carlo}. Both are exercised in the
benchmark above.

Combined, these changes do exactly what the reviewer requested:
they remove ambiguity about what the runtime numbers include, they
characterize the regime in which asymptotic MK is statistically
reliable, and they let users obtain a more principled CI on
aggregated $\alpha$ when bin counts are sparse.
}

%---------- R1.6 ----------
\begin{point}{} % 1.6 -- Ls, Ln; omega_a, omega_na decomposition
Since the authors already manually estimate the shortest mutational
paths and weight synonymous and nonsynonymous sites following Nei and
Gojobori (1986), can they include the total number of synonymous and
nonsynonymous sites ($L_s$, $L_n$)? It should then be straightforward
to also implement $\omega_a$ and $\omega_{na}$, as decomposed in
Coronado-Zamora et al.\ (2019)
(\url{https://academic.oup.com/gbe/article/11/5/1463/5480667}):
\[
\omega_a = \alpha \cdot \omega, \qquad
\omega_{na} = (1 - \alpha) \cdot \omega.
\]
\end{point}

\reply{
We agree the Nei--Gojobori site totals ($L_n$, $L_s$) are a natural
by-product of how \mkado{} already classifies codons, and the
$\omega_a$ / $\omega_{na}$ decomposition follows directly. We have
exposed both:

\begin{itemize}
\item Standard MK output now reports $L_n$, $L_s$, and
$\omega = (D_n / L_n) / (D_s / L_s)$ in addition to the usual $D_n$,
$D_s$, $P_n$, $P_s$ counts and $\alpha$, both for single-gene
\texttt{mkado test} runs and for the per-gene rows produced by
\texttt{mkado batch}.
\item Aggregated tests that yield an $\alpha$ with a confidence
interval (asymptotic MK, $\alpha_\text{TG}$) additionally report
$\omega_a = \alpha \cdot \omega$ and
$\omega_{na} = (1 - \alpha) \cdot \omega$ with their corresponding
95\% CIs propagated from the $\alpha$ CI, as decomposed by
Coronado-Zamora et al.\ (2019).
\end{itemize}

These quantities appear in both the human-readable summary and the
TSV/JSON outputs, so they can be aggregated downstream without any
custom parsing.
}

%---------- R1.7 ----------
\begin{point}{} % 1.7 -- OrthoMaM polymorphism projection writeup
The authors followed an intelligent strategy using the OrthoMaM
database to obtain human $D_n$, $D_s$, $P_n$, and $P_s$ counts. I
think projecting the polymorphic data onto the ortholog alignments
could provide an useful feature when working with non-model species
available within OrthoMaM database. I understand that including such
a feature will be challenging, but a more detailed explanation of the
polymorphism projection procedure would help users working with
non-model organisms.
\end{point}

\reply{
We agree that this projection procedure is generally useful for
non-model-species applications and have therefore documented it in
two places.

In the manuscript, the Application section now contains an
expanded methods paragraph \revref{} that walks through the
projection pipeline step by step:
(i) reconstruction of the canonical Ensembl CDS from the GRCh38
reference and the GTF-derived exon set;
(ii) matching of the ungapped \texttt{OrthoMaM} human row against
this Ensembl CDS to obtain the alignment-column to CDS-position
correspondence (with genes that fail to match dropped);
(iii) extraction of the YRI variants overlapping the
alignment-retained CDS positions, including reference-allele
verification and minus-strand allele handling;
(iv) assembly of per-individual phased haplotype CDS sequences;
and
(v) re-gapping of the haplotypes to the \texttt{OrthoMaM}
alignment, dropping CDS positions absent from the alignment and
preserving its insertion/deletion structure.
We also state explicitly that multi-transcript genes are reduced
to the canonical CDS chosen by \texttt{OrthoMaM} curation;
alternative transcripts are not analysed separately.

In the documentation, we have added a new ``Working with Ortholog
Alignments'' section to the tutorial that generalizes this
procedure to any codon-level ortholog alignment plus a
coordinate-resolved VCF, and points readers at the
\texttt{data/build\_mk\_dataset.py} script we used for the
human benchmark as a starting point for adapting it to other
species. As the reviewer kindly anticipated, we have not added a
standalone \texttt{mkado project} subcommand for this revision: the
projection procedure is sufficiently dataset-specific (alignment
source, transcript curation, VCF liftover) that a one-size-fits-all
implementation would either oversimplify the projection or impose
heavy assumptions about input formats. We are happy to revisit
this in a future release if user demand suggests a clearly
factorable subset of the procedure.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\reviewersection{2}

\begin{itquote}
This is a concise manuscript reporting a standalone package
specifically designed for computing several flavours of the classical
McDonald-Kreitman test. The length of the manuscript, level of details
and information provided are appropriate. I agree with the authors
that such as tool can be a benefit for the community by filling a gap
in the set of available programs. I would say that \mkado fulfils its
tasks successfully. I had the chance to test the package and I found
it satisfying so my review will be positive. Nevertheless I will seize
this opportunity to provide feedback after looking at the code and
testing the tool.
\end{itquote}

\reply{We thank the reviewer for the positive overall assessment and
for testing the package end-to-end. The detailed, code-level feedback
was particularly valuable; each point is addressed below.}

%---------- R2.1 ----------
\begin{point}{} % 2.1 -- formal definition of Pn and Ps
[The text has a good structure.] However, I would recommend that both
the article and the documentation describe formally the basic
statistics used. There are detailed formulas for variants of the MK
tests, $D_n$ and $D_s$ are clearly defined, but $P_n$ and $P_s$ are
not, actually (or not completely).
\end{point}

\reply{
We agree and have substantially expanded the definitions of $P_n$
and $P_s$ in the Methods \revref. The revised text now states
explicitly that $D_n$ and $D_s$ are the numbers of nonsynonymous
and synonymous fixed differences between ingroup and outgroup
under the supplied genetic code, and that $P_n$ and $P_s$ are the
corresponding numbers of ingroup-polymorphic sites at which the
derived (non-outgroup) allele segregates above an optional
frequency threshold. The revised text also makes explicit (i) the
default behaviour (no frequency filter, all polymorphic sites
counted), (ii) the effect of the \texttt{-{}-min-freq~$X$} flag,
(iii) the effect of \texttt{-{}-no-singletons} (threshold set to
$1/n$ for sample size $n$), and (iv) the handling of sites that
also segregate in the outgroup (excluded by default per the
\texttt{DnaSP} convention; included via
\texttt{-{}-pool-polymorphisms}).
We have also removed the previous standalone paragraph on
polymorphism filtering, since these details are now incorporated
directly into the formal definitions.
}

%---------- R2.2 ----------
\begin{point}{} % 2.2 -- VCF support not mentioned in manuscript
I am surprised that the possibility of importing VCF data is not
mentioned in the manuscript. I suppose that this was a recent addition
to the package. This would be an attractive addition to the article.
\end{point}

\reply{
We agree---this is a recent addition that warranted a clearer mention
in the manuscript. The same change made in response to Reviewer~1,
point~1 covers this point: the Methods now describes the
\texttt{mkado vcf} subcommand and its required inputs \revref, and
the feature comparison table (Table~\ref{tab:comparison}) gains a
dedicated ``VCF input'' row.
}

%---------- R2.3 ----------
\begin{point}{} % 2.3 -- Figure 1 caption: two-panel
Figure 1 should be cited slightly differently as it shows both the
volcano plot and the asymptotic MK test display.
\end{point}

\reply{
We agree, the in-text references and the caption were inconsistent.
The figure caption already describes the two panels with ``Left:''
and ``Right:'' labels. We have updated the four in-text references to
Figures~\ref{fig:volcano} and~\ref{fig:chimp_mk} \revref to use the
same ``left panel of'' / ``right panel of'' wording instead of the
nonexistent panel labels~A and~B; the references now make clear that
each figure shows two complementary results (volcano plot of the
per-gene standard MK test alongside the aggregated asymptotic
$\alpha(x)$ curve).
}

%---------- R2.4 ----------
\begin{point}{} % 2.4 -- __init__.py outdated (version, package name)
The main \texttt{\_\_init\_\_.py} file is outdated as of this writing.
First, the hardcoded version number is still 0.3.0 while the package
has been ported to 0.4.0. Second, the package name in the file
docstring is still ``mikado'', which is the original name of the
project.
\end{point}

\reply{
Both items are now fixed. \texttt{src/mkado/\_\_init\_\_.py} no
longer hardcodes a version string; the package version is read at
import time via \texttt{importlib.metadata.version("mkado")}, with
\texttt{pyproject.toml} as the single source of truth, so
\texttt{mkado.\_\_version\_\_} cannot drift from the released
version again. The module docstring has been updated to ``MKado:
Modern Python implementation of the McDonald-Kreitman test
toolkit.''
}

%---------- R2.5 ----------
\begin{point}{} % 2.5 -- shell completion options unclear; slow
I noticed that the umbrella command has options to manage shell
completion. They were not clear to me at first sight and it is rather
unexpected to find them as the option of the umbrella command, which
is not expected to do anything. Maybe a ``setup'' command would
clarify it but this is not essential. In any case I recommend adding a
sentence to the help page to explain that these options allow to
configure automatic shell completion. On my system, shell completion
was painfully slow so the feature was doing more harm than good.
\end{point}

\reply{
We agree that the umbrella command's behaviour should be
self-explanatory. The umbrella \texttt{mkado} help page now includes
a dedicated ``Shell completion'' note explaining what
\texttt{-{}-install-completion} and \texttt{-{}-show-completion} do
and on which shells they are supported (bash, zsh, fish,
powershell). The user-facing documentation has likewise gained a
new ``Shell Completion'' subsection in the installation page that
walks through both options, recommends a shell restart after
installing, and explicitly notes the completion can be slow on some
systems and can simply be left uninstalled---which we hope addresses
the reviewer's concrete experience without removing a feature
others may find useful.
}

%---------- R2.6 ----------
\begin{point}{} % 2.6 -- subcommand-without-args behaviour
When called without argument, the different commands end up with an
error, which is somewhat inconsistent with the behaviour of the
umbrella command (which displays the help page). But it is a perfectly
acceptable behaviour and up to the authors. Commands have logical and
relevant options.
\end{point}

\reply{
We thank the reviewer for the kind allowance. We have kept the
existing behaviour: \texttt{mkado} alone (the umbrella command)
prints the help page, while a subcommand without its required
positional arguments raises a usage error. The asymmetry is
intentional---the umbrella command does no work itself, while each
subcommand has required inputs whose absence we want to flag
explicitly with the standard non-zero exit code so that the error
is detectable in scripted workflows. The Typer-generated usage
hint produced on missing arguments points the user at the
subcommand's \texttt{-{}-help} for full documentation.
}

%---------- R2.7 ----------
\begin{point}{} % 2.7 -- output filename specification
The commands write a progress bar and some feedback to the standard
error stream, obviously to enable standard output redirection to
capture the output. Unless I missed it, there is no way to specify
[an] output file name. The choice that has been made is not perfectly
complying with conventions regarding the use of the standard error
stream but it is not unusual either.
\end{point}

\reply{
The reviewer is correct that an explicit output-file option was
missing. We have added \texttt{-{}-output / -O} to all subcommands
that produce results (\texttt{test}, \texttt{batch}, \texttt{vcf}),
which writes the formatted output to a named file rather than to
stdout. The \texttt{-{}-format} flag continues to control whether
the output is the human-readable summary, TSV, or JSON. The
progress bar and feedback messages remain on stderr, as the
reviewer correctly identified.
}

%---------- R2.8 ----------
\begin{point}{} % 2.8 -- ruff out of user manual; "third outgroup"
The package is adequately documented by a fairly complete README
available on the git and PyPI repositories, and, more in more details
by a readthedocs page. [A] small remark would be that I don't think
it is relevant to explain how to use ruff in the manual, as this is
not needed to use \mkado. By contrast, the shell completion options
could be mentioned. There is somewhere the term ``third outgroup''
which I think is incorrect. Apart for this, the documentation is
clear and complete, with very helpful commands.
\end{point}

\reply{
We have actioned all three items from this comment:

\begin{itemize}
\item The ``Linting and Formatting'' subsection has been removed
from the user-facing installation page (\texttt{docs/installation.rst}),
since \texttt{ruff} is a developer concern, not a user concern. The
\texttt{README.md} retains a brief ``Development'' section covering
\texttt{ruff} for contributors.
\item The shell completion options are now documented in the user
manual---a new ``Shell Completion'' subsection in the installation
page describes both \texttt{-{}-install-completion} and
\texttt{-{}-show-completion} (see also R2.5).
\item ``Third outgroup'' has been corrected to ``second outgroup''
in \texttt{docs/index.rst}.
\end{itemize}
}

%---------- R2.9 ----------
\begin{point}{} % 2.9 -- aggregate option meaning + range/regex matchers
Only the meaning of the ``aggregate'' option was not clear to me. As
samples are often grouped by population, it could be a possibility to
allow specifying the ingroup and outgroup by range indexes, in top of
the two possibilities already provided. This might be a feature
request. Another suggestion is to support regular expressions to
identify ingroup and outgroup samples in addition to an exact
substring match.
\end{point}

\reply{
For the \texttt{--aggregate} clarity issue, we have added a new
``Per-gene vs Aggregated Modes'' subsection to the Batch Processing
Workflow documentation that contrasts the two modes, names the
default for each test, and explains \emph{why} aggregation is the
default for asymptotic MK (per-gene SFS data are too sparse to fit).

For the two new selector forms, we have decided to decline both.
Sample names in the formats \mkado{} accepts (multi-sample VCF,
multi-record FASTA) typically carry the relevant grouping---1000
Genomes population codes, OrthoMaM species tags, ad-hoc user
prefixes---so substring matching already covers the
population-grouped case. Substring is also \emph{order-invariant}: a
VCF re-header or FASTA shuffle leaves a name-based filter
unchanged, while an index-range filter silently selects different
samples. Adding ranges would also give \texttt{-i} and \texttt{-o}
two semantics that every consumer (CLI, batch workers, VCF
subcommand, Python API) would have to dispatch on, and we expect
the format would accrete (unions, open-ended ranges, slice syntax)
into a small sublanguage. Regex matchers compound this with their
classic unanchored-pattern footgun---silently selecting the wrong
samples is the worst class of failure for a selection-inference
tool. Substring's clean ``no matches found'' is a narrower
failure mode, and the remaining gap is small enough not to
justify the risk. If a case arises where neither substring nor an
explicit samples-list file suffices, we will revisit.
}

%---------- R2.10 ----------
\begin{point}{} % 2.10 -- Figure S1 linear scaling claim
For large datasets it is still needed to use parallelization. However,
Python isn't (currently) really good at parallelization. My feeling is
that processing one alignment doesn't involve a lot of computation
compared with the cost of spawning a worker (which entails creating a
whole Python process) so the number of workers should be orders of
magnitude below the number of genes (to avoid a waste of resources for
little benefit). Figure S1 actually shows that the relationship is all
but linear so I am not sure to agree with the sentence claiming that
speedup is linear until 32 workers.
\end{point}

\reply{
The reviewer is correct: the claim of ``near-linear scaling up to
32 threads'' was not well supported by the figure. We have rewritten
the relevant paragraph in the Methods \revref to reflect what the
figure actually shows. The new text states that scaling is
approximately linear only up to roughly~16 workers, with
progressively diminishing efficiency beyond~32, and explains that
this is a direct consequence of per-gene work (parsing, codon
classification, and the per-gene MK test) being on the order of
milliseconds while the spawn cost of a Python \texttt{multiprocessing}
worker is fixed. We now explicitly recommend setting the worker
count well below the gene count for batch runs at any scale.
The Figure~S1 caption has been amended to make the same point.

We have also added a complementary supplementary figure
(\cref{fig:aggregated_runtime}) showing scaling for an
aggregated-asymptotic workload, where per-task work is much larger
and the scaling regime is therefore quite different. This is
discussed in our response to Reviewer~1, point~5.
}
